My wonderful paper

### 2024-12-06 _manuscript/_tex/index.tex
author
rewrite Section 3.1

\thanks{CONTACT: H. Sherry
Zhang. Email: \href{mailto:huize.zhang@austin.utexas.edu}{\nolinkurl{huize.zhang@austin.utexas.edu}}. Roger
D.
Peng. Email: \href{mailto:roger.peng@austin.utexas.edu}{\nolinkurl{roger.peng@austin.utexas.edu}}. }
\begin{document}
\captionsetup{labelsep=space}
\maketitle
\textsuperscript{1}  University of Texas at Austin,  
\begin{abstract}
Unexpected results in data analysis can prompt researchers to examine
data quality and analysis steps. While some researchers can typically
diagnose issues effectively with a few checks, others may struggle to
identify appropriate checks to diagnose the problem. {[}an example of
check{]} These checks are often informal and difficult to trace and
discuss in publications, resulting in others questioning the
trustworthiness of the analysis. To address this, we formalize the
informal checks into an \emph{analysis plan} that encompasses the
analysis steps and (the unit tests): one test for whether the result
meets expectations and multiple tests for checking the analysis. We then
present a procedure to assess the quality of these unit tests based on
their accuracy and redundancy on simulated versions of the original
data. The accuracy is assessed using binary classification metrics,
\emph{i.e.}, precision and recall, while redundancy is calculated using
mutual information. This procedure can be used to conduct a sensitivity
analysis, compare different analysis plans, and to identify the optimal
cutoff point for the unit tests.
\end{abstract}

@@

In data analysis, experienced researchers often rely on their prior
knowledge or domain expertise to quickly assess whether results align
with their expectations. When outcomes deviate, they can quickly iterate
through the analysis cycle to identify potential issues, refine methods,
or revise assumptions. However, this iterative process they employ is
rarely explicit, documented, or discussed in publications. As a result,
readers are typically presented with the final outcomes of the analysis
cycle where the results and expectations are aligned -- achieved either
by refining the analysis or updating the expectations based on
statistical evidence. These missing pieces of information provides
little guidance for diagnosing issues in the analysis when the same
methodology is applied to a new dataset that produces different
outcomes. Similarly, when researchers with different background
knowledge it becomes unclear whether variations in results arise from
differences in the assumptions or in the analysis itself.

@@

@@

A substantial body of literature has addressed the definition of data
quality \citep[more]{8642813} and developed frameworks, that includes
dimensions, attributes, and measures to evaluate and improve data
quality
\citep{cai2015challenges, wang1996beyond, 6204995, woodall2014classification}.
These frameworks are often used information system and database
management and support business decision-making in the industry. For
research purposes, high-quality data ensures the credibility of
scientific findings and supports reproducibility and reusability in
future studies {[}ref{]}. With the growing prevalence of open data in
scientific research, the consumers or users of the data typically are no
longer the data producers or collectors who have the full knowledge of
data in hand, prompting more interest towards the data quality checks in
the data analysis process. In R, there are some packages,
like\texttt{skimr} \citep{skimr} and \texttt{dataMaid} \citep{dataMaid},
provides basic data screening and reporting, while another class of
packages, e.g.~\texttt{assertr} \citep{assertr}, \texttt{validate}
\citep{validate}, and \texttt{pointblank} \citep{pointblank}, focuses on
providing data validation tools, allowing users to define customized
data quality checks based on the applications.

@@

Q: Whether we should formulate these concept with math notation?? A:
only if it helps

While data validation focuses on verifying or subsetting data based on
predefined rules \citep{zio2015methodology}, data analysis checks
specify assumptions made on the data required for the analysis process
to obtain the reported results. These checks often involve evaluating
aspects such as data distribution and outliers, relationships between
variables, and the validity of model assumptions.

However, there are other checks that reflect the mental process of of
how data analysts diagnose the unexpected output that are not documented
in this process. For example, when an unexpected output occurs, an
analyst may check on whether a column in the data frame is between two
values for data quality. However, it is not documented what motivates
the analyst to conduct this check -- whether it also applies to other
researchers analyzing new data in the same contexts, whether it is a
common practice in the field, or whether it is a reaction to this
particular data or scenario. Currently, most of these assumptions are
not discussed in the publication or captured by tools themselves. While
one might be able to infer some of the mental process of the analysts
from external sources, such as talking to them or watching screencast
videos produced by the analysts e.g.~TidyTuesday screencast videos,
these insights are not systematically documented or made
machine-readable. A question to consider here is whether it is possible
to deliver this type of quality checks as structured data in a data
analysis. This gap highlights the need for tools that provide higher
level documentation of reasoning behind the checks, facilitating a more
transparent and interpretable analysis process.

An analysis plan is a set of analysis steps combined with expectations.
Expectations represent our belief about certain aspects of the analysis,
independent of the analysis itself. It can be divided into two types:
\emph{outcome expectation} and \emph{plan expectation}. Outcome
expectation refers to what we anticipate from the main result of the
analysis based on prior knowledge. They shape how we interpret the
results and assess whether they are consistent with existing knowledge
or indicate the need for updates \citep{grolemund_cognitive_2014}. For
example, in public health, prior research shows the average increase in
mortality rate per unit increase in PM10 is about 0.50\%
\citep{liu2019ambient}. This serves as an expectation for similar future
studies. Plan expectations concern the intermediate steps within the
analysis rather than the final outcome. They serve as checkpoints to
detect deflection in the analysis process For example, we may expect
temporal data to be ordered with no gaps and duplicates, or expect that
temperature will be a significant covariate in the linear regression
model of PM10 on mortality.

Experienced analysts often have implicit expectation about the outcome
and rely on a few ``directional signs'' to check when the outcome
deviate from those expectation. However, these expectations are rarely
made explicit within the analysis workflow. This makes it challenging
for consumers of the analysis to evaluate the results, since it becomes
difficult to disentangle whether discrepancies arise from differing
expectations or from the use of statistical technique, without running
the analysis themselves. Non-expert analysts, lacking prior knowledge or
instinct, may not have clear expectations of the results. This can lead
to reduced confidence of the analysis and makes it more difficult and
time-consuming to diagnose the cause of the deviation when the results
don't align with expectations. By explicitly formulating these
expectations, an analysis plan can guide the analysis process,
facilitate the communication and evaluate the validity of the results.

Yet, making them explicit is crucial for

\begin{enumerate}
\def\labelenumi{\arabic{enumi})}
\tightlist
\item
  helping junior researchers interpret results and diagnose the
  unexpected,
\item
  providing checkpoints that support independent replication or
  application of methods to new data, and
\item
  aligning assumption across researchers from different backgrounds.
\end{enumerate}

>>>>

\thanks{CONTACT: H. Sherry
Zhang. Email: \href{mailto:huize.zhang@austin.utexas.edu}{\nolinkurl{huize.zhang@austin.utexas.edu}}. Roger
D.
Peng. Email: \href{mailto:roger.peng@austin.utexas.edu}{\nolinkurl{roger.peng@austin.utexas.edu}}. }
\begin{document}
\captionsetup{labelsep=space}
\maketitle
\textsuperscript{1} Department of Statistics and Data
Sciences, University of Texas at Austin, Texas, United States
\begin{abstract}
Unexpected results in data analysis can prompt researchers to examine
data quality and analysis steps. While some researchers can typically
diagnose issues effectively with a few checks, others may struggle to
identify appropriate checks to diagnose the problem. {[}an example of
check{]} These checks are often informal and difficult to trace and
discuss in publications, resulting in others questioning the
trustworthiness of the analysis. To address this, we formalize the
informal checks into an \emph{analysis plan} that encompasses the
analysis steps and (the unit tests): one test for whether the result
meets expectations and multiple tests for checking the analysis. We then
present a procedure to assess the quality of these unit tests based on
their accuracy and redundancy on simulated versions of the original
data. The accuracy is assessed using binary classification metrics,
\emph{i.e.}, precision and recall, while redundancy is calculated using
mutual information. This procedure can be used to conduct a sensitivity
analysis, compare different analysis plans, and to identify the optimal
cutoff point for the unit tests.
\end{abstract}

@@

In data analysis, experienced researchers often rely on their prior
knowledge or domain expertise to quickly assess whether results align
with their expectations. When a result falls outside of this interval,
it prompts the researchers to investigate backwards on the data quality,
the analysis steps, or the assumptions made during the analysis process.
This mental process of where to diagnose unexpected outcomes is often
difficult to trace and discuss in publications. As a result, readers are
typically presented with the final outcomes of the analysis cycle where
the results and expectations are aligned -- achieved either by refining
the analysis or updating the expectations based on statistical evidence
\citep{grolemund_cognitive_2014}. These missing pieces of information
provides little guidance for diagnosing issues in the analysis when the
same methodology is applied to a new dataset that produces different
outcomes. Similarly, when researchers with different background
knowledge it becomes unclear whether discrepancies arise from differing
expectations or from the use of statistical techniques.

One might gain insight into analysts' thought process by speaking with
them directly or watching screencast videos they produce, such as,
TidyTuesday screencast videos. However, direct conversation are not
scalable and may not be feasible while creating educational videos
requires significant effort from the researchers. Ideally, there could
be a way to make these expectations explicit and accessible to others.
Even better, if the encoding is machine-readable, we could analyze these
expectations and learn from the analysis -- whether the checks also
apply to other researchers analyzing new data in the same context,
whether they reflect commo common practices in the field, or whether
they are sepcific to the data or analysis at hand.

@@

They shape how we interpret the results and assess whether they are
consistent with existing knowledge or indicate the need for updates.
\citep{grolemund_cognitive_2014}

@@

A substantial body of literature has addressed the definition of data
quality \citep[more]{8642813} and developed frameworks, that includes
dimensions, attributes, and measures to evaluate and improve data
quality
\citep{cai2015challenges, wang1996beyond, 6204995, woodall2014classification}.
These frameworks are often used information system and database
management and support business decision-making in the industry. For
research purposes, high-quality data ensures the credibility of
scientific findings and supports reproducibility and reusability in
future studies {[}ref{]}. With the growing prevalence of open data in
scientific research, the consumers or users of the data typically are no
longer the data producers or collectors who have the full knowledge of
data in hand, prompting more interest towards the data quality checks in
the data analysis process. In R, there are some packages, like
\texttt{skimr} \citep{skimr} and \texttt{dataMaid} \citep{dataMaid},
provides basic data screening and reporting, while another class of
packages, e.g.~\texttt{assertr} \citep{assertr}, \texttt{validate}
\citep{validate}, and \texttt{pointblank} \citep{pointblank} focuses on
providing data validation tools, allowing users to define customized
data quality checks based on the applications.

@@

Expectations represent our beliefs about certain aspects of the
analysis, independent of the analysis itself. When outcomes deviate from
these expectations, analysts often revisit the analysis cycle to
identify potential issues, refine methods, or revise assumptions.
Experienced analysts can typically identify issues quickly and correct
them on the spot, but they often do so without discussing the underlying
reasoning, making it harder for less experienced researchers to learn
and master these skills. Here, we introduce the concept of
\emph{analysis validation checks}, which frames these expectations or
assumptions as explicit checks that returns a TRUE or FALSE output given
the data analyzed at hand. Inspired by \emph{data validation checks}
\citep{validate}, where ensure datasets meet expected format and
quality, analysis validation checks reverse the approach: they validate
the assumptions about the data necessary for the analysis to produce the
reported results. This allows the concept to encompass a broad range of
checks, such as data quality (i.e.~missing data, how the data is
structured), data distribution and outliers, bivariate and multivariate
relationships between variables, and contextual information. These
externalized checks provide insights into analysts' thoughts process
process and offer several benefits:

\begin{enumerate}
\def\labelenumi{\arabic{enumi})}
\tightlist
\item
  help students and junior researchers develop skills to diagnose
  unexpected analysis results,
\item
  serve as clear checkpoints to support independent replication or
  application of methods to new data, and
\item
  align assumptions among researchers from different domain backgrounds.
\item
  improve analysis transparency, reproducibility, and trustworthiness.
\end{enumerate}

### 2024-12-06 index.qmd
rewrite Section 3.1

In data analysis, experienced researchers often rely on their prior knowledge or domain expertise to quickly assess whether results align with their expectations. When outcomes deviate, they can quickly iterate through the analysis cycle to identify potential issues, refine methods, or revise assumptions. However, this iterative process they employ is rarely explicit, documented, or discussed in publications. As a result, readers are typically presented with the final outcomes of the analysis cycle where the results and expectations are aligned -- achieved either by refining the analysis or updating the expectations based on statistical evidence. These missing pieces of information provides little guidance for diagnosing issues in the analysis when the same methodology is applied to a new dataset that produces different outcomes. Similarly, when researchers with different background knowledge it becomes unclear whether variations in results arise from differences in the assumptions or in the analysis itself.

@@

@@

A substantial body of literature has addressed the definition of data quality [@8642813, more] and developed frameworks, that includes dimensions, attributes, and measures to evaluate and improve data quality [@cai2015challenges; @wang1996beyond; @6204995; @woodall2014classification]. These frameworks are often used information system and database management and support business decision-making in the industry. For research purposes, high-quality data ensures the credibility of scientific findings and supports reproducibility and reusability in future studies [ref]. With the growing prevalence of open data in scientific research, the consumers or users of the data typically are no longer the data producers or collectors who have the full knowledge of data in hand, prompting more interest towards the data quality checks in the data analysis process. In R, there are some packages, like`skimr` [@skimr] and `dataMaid` [@dataMaid], provides basic data screening and reporting, while another class of packages, e.g. `assertr` [@assertr], `validate` [@validate], and `pointblank` [@pointblank], focuses on providing data validation tools, allowing users to define customized data quality checks based on the applications.

@@

Q: Whether we should formulate these concept with math notation?? A: only if it helps

While data validation focuses on verifying or subsetting data based on predefined rules [@zio2015methodology], data analysis checks specify assumptions made on the data required for the analysis process to obtain the reported results. These checks often involve evaluating aspects such as data distribution and outliers, relationships between variables, and the validity of model assumptions.

However, there are other checks that reflect the mental process of of how data analysts diagnose the unexpected output that are not documented in this process. For example, when an unexpected output occurs, an analyst may check on whether a column in the data frame is between two values for data quality. However, it is not documented what motivates the analyst to conduct this check -- whether it also applies to other researchers analyzing new data in the same contexts, whether it is a common practice in the field, or whether it is a reaction to this particular data or scenario. Currently, most of these assumptions are not discussed in the publication or captured by tools themselves. While one might be able to infer some of the mental process of the analysts from external sources, such as talking to them or watching screencast videos produced by the analysts e.g. TidyTuesday screencast videos, these insights are not systematically documented or made machine-readable. A question to consider here is whether it is possible to deliver this type of quality checks as structured data in a data analysis. This gap highlights the need for tools that provide higher level documentation of reasoning behind the checks, facilitating a more transparent and interpretable analysis process.

An analysis plan is a set of analysis steps combined with expectations. Expectations represent our belief about certain aspects of the analysis, independent of the analysis itself. It can be divided into two types: *outcome expectation* and *plan expectation*. Outcome expectation refers to what we anticipate from the main result of the analysis based on prior knowledge. They shape how we interpret the results and assess whether they are consistent with existing knowledge or indicate the need for updates [@grolemund_cognitive_2014]. For example, in public health, prior research shows the average increase in mortality rate per unit increase in PM10 is about 0.50% [@liu2019ambient]. This serves as an expectation for similar future studies. Plan expectations concern the intermediate steps within the analysis rather than the final outcome. They serve as checkpoints to detect deflection in the analysis process For example, we may expect temporal data to be ordered with no gaps and duplicates, or expect that temperature will be a significant covariate in the linear regression model of PM10 on mortality.

Experienced analysts often have implicit expectation about the outcome and rely on a few "directional signs" to check when the outcome deviate from those expectation. However, these expectations are rarely made explicit within the analysis workflow. This makes it challenging for consumers of the analysis to evaluate the results, since it becomes difficult to disentangle whether discrepancies arise from differing expectations or from the use of statistical technique, without running the analysis themselves. Non-expert analysts, lacking prior knowledge or instinct, may not have clear expectations of the results. This can lead to reduced confidence of the analysis and makes it more difficult and time-consuming to diagnose the cause of the deviation when the results don't align with expectations. By explicitly formulating these expectations, an analysis plan can guide the analysis process, facilitate the communication and evaluate the validity of the results.

Yet, making them explicit is crucial for

  1) helping junior researchers interpret results and diagnose the unexpected, 
  2) providing checkpoints that support independent replication or application of methods to new data, and
  3) aligning assumption across researchers from different backgrounds.

<!-- The expectations can be thought of as a set of unit tests used to validate the results of data analysis. By specifying a range of values for these tests, multiple versions of the dataset can be generated to satisfy different sets of plan expectations. This allows us to present what we called the "result universe" -- the complete set of possible results that can be obtained from one data analysis process. By visualizing the result universe, data analysis consumers can observe how changes in expectations affect the results and the range of alternative outcomes that could arise under different conditions. This enables them to evaluate the outcomes based on their own plan expectations and gain a broader perspective on how the actual results produced by analysts fit within this spectrum of possibilities, promoting transparency and trust in analysis.   -->

>>>>

In data analysis, experienced researchers often rely on their prior knowledge or domain expertise to quickly assess whether results align with their expectations. When a result falls outside of this interval, it prompts the researchers to investigate backwards on the data quality, the analysis steps, or the assumptions made during the analysis process. This mental process of where to diagnose unexpected outcomes is often difficult to trace and discuss in publications. As a result, readers are typically presented with the final outcomes of the analysis cycle where the results and expectations are aligned -- achieved either by refining the analysis or updating the expectations based on statistical evidence [@grolemund_cognitive_2014]. These missing pieces of information provides little guidance for diagnosing issues in the analysis when the same methodology is applied to a new dataset that produces different outcomes. Similarly, when researchers with different background knowledge it becomes unclear whether discrepancies arise from differing expectations or from the use of statistical techniques.

One might gain insight into analysts' thought process by speaking with them directly or watching screencast videos they produce, such as, TidyTuesday screencast videos. However, direct conversation are not scalable and may not be feasible while creating educational videos requires significant effort from the researchers. Ideally, there could be a way to make these expectations explicit and accessible to others. Even better, if the encoding is machine-readable, we could analyze these expectations and learn from the analysis -- whether the checks also apply to other researchers analyzing new data in the same context, whether they reflect commo common practices in the field, or whether they are sepcific to the data or analysis at hand.

@@

They shape how we interpret the results and assess whether they are consistent with existing knowledge or indicate the need for updates. [@grolemund_cognitive_2014]

@@

A substantial body of literature has addressed the definition of data quality [@8642813, more] and developed frameworks, that includes dimensions, attributes, and measures to evaluate and improve data quality [@cai2015challenges; @wang1996beyond; @6204995; @woodall2014classification]. These frameworks are often used information system and database management and support business decision-making in the industry. For research purposes, high-quality data ensures the credibility of scientific findings and supports reproducibility and reusability in future studies [ref]. With the growing prevalence of open data in scientific research, the consumers or users of the data typically are no longer the data producers or collectors who have the full knowledge of data in hand, prompting more interest towards the data quality checks in the data analysis process. In R, there are some packages, like `skimr` [@skimr] and `dataMaid` [@dataMaid], provides basic data screening and reporting, while another class of packages, e.g. `assertr` [@assertr], `validate` [@validate], and `pointblank` [@pointblank] focuses on providing data validation tools, allowing users to define customized data quality checks based on the applications.

@@

<!-- We may anticipate certain interval on the main result of the analysis based on prior knowledge. For example, researchers in public health may expect an average increase in mortality rate per unit increase in PM10 is about 0.50% [@liu2019ambient] [TODO: check if this is correct].  -->

Expectations represent our beliefs about certain aspects of the analysis, independent of the analysis itself. When outcomes deviate from these expectations, analysts often revisit the analysis cycle to identify potential issues, refine methods, or revise assumptions. Experienced analysts can typically identify issues quickly and correct them on the spot, but they often do so without discussing the underlying reasoning, making it harder for less experienced researchers to learn and master these skills. Here, we introduce the concept of *analysis validation checks*, which frames these expectations or assumptions as explicit checks that returns a TRUE or FALSE output given the data analyzed at hand. Inspired by *data validation checks* [ @validate], where ensure datasets meet expected format and quality, analysis validation checks reverse the approach: they validate the assumptions about the data necessary for the analysis to produce the reported results. This allows the concept to encompass a broad range of checks, such as data quality (i.e. missing data, how the data is structured), data distribution and outliers, bivariate and multivariate relationships between variables, and contextual information. These externalized checks provide insights into analysts' thoughts process process and offer several benefits:

  1) help students and junior researchers develop skills to diagnose unexpected analysis results, 
  2) serve as clear checkpoints to support independent replication or application of methods to new data, and 
  3) align assumptions among researchers from different domain backgrounds.
  4) improve analysis transparency, reproducibility, and trustworthiness.
